---
title: "human_gest_cardiac_exprs"
author: "Juan C. Sanchez-Arias"
date: "31/03/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Set up

Load necessary packages for subsequent analysis

```{r other-packages, include=TRUE, warning=FALSE}

#General Bioconductor packages
    library(Biobase)
    library(oligoClasses)
     
#Annotation and data import packages
    library(ArrayExpress)
    library(pd.hugene.1.0.st.v1)
    library(hugene10sttranscriptcluster.db)
     
#Quality control and pre-processing packages
    library(oligo)
    library(arrayQualityMetrics)
     
#Analysis and statistics packages
    library(limma)
    library(topGO)
    library(DESeq2)
    library(ReactomePA)
    library(clusterProfiler)
     
#Plotting and color options packages
    library(gplots)
    library(ggplot2)
    library(geneplotter)
    library(RColorBrewer)
    library(pheatmap)
    library(enrichplot)
     
#Formatting/documentation packages
   #library(rmarkdown)
   #library(BiocStyle)
    library(dplyr)
    library(tidyr)

#Helpers:
    library(stringr)
    library(matrixStats)
    library(genefilter)
    library(openxlsx)

```

## 2. Downloading data from ArrayExpress

```{r raw-data-check, echo=TRUE}

data_dir <- tempdir()

if (!dir.exists(data_dir)) {
    dir.create(data_dir)
}
```

The function `getAE` downloads the data into the temporary directory (`data_dir`).

```{r raw-data-import, include=TRUE, message=FALSE, warning=FALSE}

anno_AE <- getAE("E-MTAB-7031", path = data_dir)
```

Each dataset in ArrayExpress is stored following MAGE-TAB (MicroArray Gene Expression Tabular) specifications:

-   Investigation Description Format (IDF): contains top level information about the experiment including title, description, submitter contact details and protocols
-   Array Design Format (ADF)
-   Sample and Data Relationship Format (SDRF): contains essential information on the experimental samples, e.g. the experimental group(s) they belong to.
-   Raw data files
-   Processed data files

## 3. Creating ExpressionSet with `Biobase`

An `ExpressionSet` can be manipulated (e.g., subsetted, copied), and is the input to or output of many Bioconductor functions.

The data in an `ExpressionSet` consist of:

-   assayData: Expression data from microarray experiments with microarray probes in rows and sample identifiers in - columns
-   metaData
-   phenoData: A description of the samples in the experiment with sample identifiers in rows and description elements in columns; holds the content of the SDRF file
-   featureData: metadata about the features on the chip or technology used for the experiment with same rows as assayData by default and freely assignable columns
-   further annotations for the features, for example gene annotations from biomedical databases (annotation). experimentData: A flexible structure to describe the experiment.

> The functions `pData` and `fData` extract the sample and feature annotation, respectively, from an ExpressionSet. The function `exprs` will return the expression data itself as a matrix.

```{r SDRF, include=TRUE}

sdrf_location <- file.path(data_dir, anno_AE$sdrf)
SDRF <- read.delim(sdrf_location)

rownames(SDRF) <- SDRF$Scan.Name
# SDRF <- AnnotatedDataFrame(SDRF)
```

```{r, include=TRUE}

cts <- read.delim("E-MTAB-7031-raw-counts.tsv", row.names = 1)

gene.names <- cts$Gene.Name

cts <- cts[,-c(1)]

colID <- c("ERR2704715", "ERR2704716", "ERR2704714", "ERR2704720", "ERR2704719", "ERR2704713", "ERR2704717", "ERR2704718", "ERR2704712")

dev.stage <- c("16 WGA", "16 WGA", "14 WGA", "9 WGA", "9 WGA", "12 WGA", "16 WGA", "9WGA", "12 WGA")

coldata <- data.frame(colID, dev.stage, row.names = colnames(cts))
```

```{r}
cts_2 <- cts[, -c(3, 6, 9)]

colID_2 <- c("ERR2704715", "ERR2704716", "ERR2704720", "ERR2704719", "ERR2704717", "ERR2704718")

dev.stage_2 <- c("16 WGA", "16 WGA", "9 WGA", "9 WGA", "16 WGA", "9WGA")

coldata_2 <- data.frame(colID_2, dev.stage_2, row.names = colnames(cts_2))
```

```{r DESeqDataSet-01, include=TRUE}
dds <- DESeqDataSetFromMatrix(countData = cts,
                              colData = coldata,
                              design = ~ dev.stage)
rld <- rlog(dds, blind=FALSE)
```

```{r DESeqDataSet-01, include=TRUE}
dds_2 <- DESeqDataSetFromMatrix(countData = cts_2,
                              colData = coldata_2,
                              design = ~ dev.stage_2)
```

```{r}
dds_2 <- DESeq(dds_2)

dds <- DESeq(dds)
```

```{r}
res <- results(dds, contrast= c("dev.stage", "16 WGA","9 WGA"))
res
```
```{r}
resOrdered <- res[order(res$pvalue),]
```

```{r write, include=TRUE}
write.csv(as.data.frame(resOrdered), 
          file="human_wga_16_9.csv")
```

```{r}
write.csv(as.data.frame(assay(rld)), 
          file="human_gest_Cardiac_expr_rld.csv")
```

